\documentclass[a4paper, twoside, openright, 11pt, oldfontcommands]{memoir}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{algorithm}
\usepackage{algpseudocode}
\OnehalfSpacing
\usepackage{cite}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{subcaption}
\usepackage{fixltx2e}
\usepackage{hyperref}
\usepackage{stmaryrd}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
\usepackage[authoryear,square]{natbib}
\usepackage{bibentry}
\renewcommand{\cite}{\citep}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,automata,shapes.multipart,positioning}

\let\proof\relax
\let\endproof\relax
\usepackage{amsthm}
\usepackage{xspace}

\algnewcommand{\IIf}[1]{\State\algorithmicif\ #1\ \algorithmicthen}
\algnewcommand{\EndIIf}{}
\algnewcommand{\IfElse}[3]{\algorithmicif\ #1\ \algorithmicthen\ #2\ \algorithmicelse\ #3}
\algnewcommand{\EndIfElse}{}
\algnewcommand{\Break}{\State\textbf{break}}
\MakeRobust{\Call}
\newcommand{\True}{\texttt{true}}
\newcommand{\False}{\texttt{false}}
\newcommand{\II}{\mathcal{I}}
\newcommand{\textoverline}[1]{$\overline{\mbox{#1}}$}

%%% Formatting setup
% 11pt Palatino (URW Palladio) for text
\renewcommand{\rmdefault}{ppl}
% Optima (URW Classico) for headings
%\renewcommand{\sfdefault}{uop}

% Use Courier which has a bold series
\renewcommand{\ttdefault}{pcr}

% Make our pretty chapter headings
% Copied from DaveC via Adam
\usepackage[bf,sf]{titlesec}
\newcommand{\chapformat}[1]{\parbox[c]{0.8\textwidth}{\Huge #1}}
\titleformat{\chapter}[hang]
    {\sffamily\bfseries}
    {\parbox[c]{0.2\textwidth}{\rmfamily\fontsize{72}{72}
     \selectfont\thechapter\hspace{10pt}\rule[-12pt]{2pt}{72pt}}}
    {0pt}
    {\chapformat}

\title{Addressing State Explosion in Reactive Synthesis}

\author{Alexander Legg \\
    \\
    Supervised by \\
Leonid Ryzhyk, Nina Narodytska, and Gernot Heiser}

\begin{document}

\maketitle

\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\tableofcontents

\newpage

\begin{abstract}

Software controllers of reactive systems are ubiquitous in situations where incorrectness has a high cost. In order to place trust in the software, strong guarantees of its functional correctness are required. Reactive synthesis can be used to automatically construct software to a specification and ensure correctness. The drawback is that synthesis is computationally hard meaning that many specifications are infeasible to synthesise.

Synthesis is hard because reactive systems have many possible states that the synthesis procedure must consider. A standard way to mitigate this complexity is to symbolically represent sets of states with data structures such as binary decision diagrams (BDDs). The usual approach to synthesis is to construct the set of all states that are safe for the controller. In some specifications this set has no compact representation as a BDD and so it is slow or impossible to synthesise a controller in this way.

In this thesis I propose a synthesis algorithm that constructs an approximation of the set of safe states that is sufficient to show correctness of the controller. The algorithm enables synthesis of specifications that were previously infeasible.

\end{abstract}

\newpage

\renewcommand{\abstractname}{Acknowledgements}
\begin{abstract}
Acknowledge some people
\end{abstract}

\newpage

\renewcommand{\abstractname}{Publications}
\begin{abstract}
List publications
\end{abstract}

\input{introduction.tex}

\input{background.tex}

\chapter{Related Work}
\label{ch:relatedwork}

\input{bounded.tex}

\input{strategy.tex}

\input{unbounded.tex}

\chapter{Evaluation}

\chapter{Conclusion}

\bibliographystyle{thesisnat}
\bibliography{main}

\end{document}
